---
title: "Data Wrangling with R Basics - 1"
date: 2020-04-24
draft: false
tags: ["R", "Data Wrangling", "tutorial"]
categories: ["English"]
author: "Yaoyu E. Wang"

autoCollapseToc: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The R syntax and Data Structre

R is composed of the following syntax to describe the program:
- *variables* that can be of any of R data types
- *operators* operations that perform a arithematic or numerical action
- *functions* a description of a series of operations to be performed
- *comments*  the **#** indicates comments that are not performed by R

## Variables and Data types
*Variables* are assignment names for a piece of information, such as data.  The data can take the form of simple *data type* such as integer, characters, or logics, or can take the form of more complex *data structure*.


|Operator|Description|Example
|:---:|:---|:---
|numeric|any number including floating| 1, 1.234
|integer|integer number| 1,2,3,4...
|character|any string of text values declared by quotation **"** or **'**|"a", "abcd", '123'
|logical|boolean results| TRUE/FALSE

There are other data types that are beyond scope of this course.

We can declare a simple variable and find its data type with the following:

```{r variable}
# This is a comment that is not read by R 
# Comment is for human to read to explain the code
x=5
x<-10
# Declare a character
y="a"
# Declare a logic
z=TRUE
```

We can also find the data type of a variable by **class** function.
```{r find_class, results='hide'}
class(x)
class(y)
class(z)
```

## Data Structure

**Variables** can be stored in different data structure that takes different types of variable organization.

### Vector
The simplest is the **vector** which contains a series of values with the same data type.  
We can also declare a vector of numbers into a single variable using **c()**

For example we can declare the following vectors with different data types:
```{r vector, include=TRUE, results="asis"}
# declare a numeric vector
x=c(1,2,3,4)
# declare a character vector
y=c("a","b","c")
# declare a logic vector, TRUE and FALSE can be short-handed as T and F
z=c(TRUE,FALSE,T,F)

# A data type is assumed when a mixed data types are declared
a=c(1,2,3,"f")
class(a)  # a would be a character vector

b=c(5, F)   # this is a numeric vector  F becomes 0
```
Since **vector** can only be of a single data type, a data type is assumed when a mixed data types of vector is declared.  The default priority is given to **character**->**numeric**/**integer**->**logic**

The data type can also be changed into different types:
```{r casting, results='hide'}
class(x)
x=as.character(x)
x
class(x)
x=as.numeric(x)
x
```

We can get the length of the vector as well as access values by calling for their specific positions using square brackets.
```{r get_values, results="asis"}
# length can be determined 
len=length(x)
len

x[1]
x[2]
```

### Matrix
The vector can be tranformed into a matrix by assign a dimension.  For instance, a vector with length of 4 can be transformed into a 2x2 matrix by **dim** function.  A matrix can also be created by **matrix** function. Again,  the data type of the matrix needs to be uniform as either numeric, character, or logical.

```{r matrix, results="markup"}
# change vector into matrix by dim
x=c(1,2,3,4)
dim(x)=c(2,2)
x
# change vector into matrix by declare matirx
y=c(1,2,3,4)
y=matrix(y, nrow=2, ncol=2)
y

# declare a character vector into a matrix
a=matrix(c('a','b', 'c', 'd'), nrow=2, ncol=2)
a
```

Values within the matrix can be access through defining value coordinates in very similar way as excel sheet with row and column numbers.
```{r matrix_value, results="markup"}
x[1,1]
x[2,1]
```




### List
A **list** is similar to vector, but allow for mixing data type of vectors to be stored in a single variable as elements with assigned name. A list variable can be declare as a vector of vectors with or without names.  The assigned values still retain its order.

For example:

```{r list, results='hide'}
# a list variable can be declare as a vector of vectors with or without names
y=list(a=1, 17, b=2:5, c='a')
```
Assigns 4 elements into **y** with the order of 1, 17, 2:5, and 'a'.  These values can be access by either name with a $ or position within the list:

```{r list value, results="markup"}
y[[1]]
y[[2]]
y$b
```

We can also get the names from the list variable and rename the elements:
```{r list_name}
names(y)
names(y)=c("a","b","c","d")
names(y)
y$b
```

### Data Frame
**Data Frame** is a special kind of list to store rectangular data sets  Think of it as excel sheet where each column is a list.  

```{r df}
df.y=data.frame(y)
df.y
```

The *data.frame* function transform a list into a rectangular data frame with the number of row is equaled to the longest element.  In this case it is column c, which contains 4 values.  All other shorter values are repeated to fit into the rectangular data frame.

We can use $ to retreive a list within DF, it is equivalent to select a column in an excel sheet. When we combine list name with a number,  it is equivalent defining a cell in excel with column name and row number like **'A1'** points to column A, row 1 cell.

```{r results="markup"}
df.y$a[1]
df.y$c[1]
df.y$c
```

### Factors
**Factor** is a data type unique to R and other statistical language, it condenses variables into unique categorical variables and store them as levels.  Makes computation of large data with repeated string/character data type more efficient

```{r}
x=sample(letters[1:5], 10, replace = TRUE)
x
y=factor(x)
y
levels(y)
as.numeric(y)
```


## Fundamental Operators

R contains designed arithmetic and logic operators:
### Arithmetic
|Operator|Description|
|:---:|:---|
|+	|addition
|-	|subtraction
|*	|multiplication
|/	|division
|^ or **	|exponentiation
|x %% y	|modulus (x mod y) 5%%2 is 1
|x %/% y	|integer division 5%/%2 is 2

### Logic
|Operator|Description|
|:---:|:---|
|<	|less than
|<=	|less than or equal to
|\>	|greater than
|\>=	|greater than or equal to
|=	|exactly equal to
|!=	|not equal to
|!x	|Not x
|x \| y	|x OR y
|x & y	|x AND y
|isTRUE(x)	|test if X is TRUE

For example we can add two matrices together using **+**
```{r}
x=matrix(c(1,2,3,4), nrow=2, ncol=2)
y=matrix(c(5,6,7,8), nrow=2, ncol=2)
x+y
```
or compare them
```{r}
x>y
```
